---
title: "03_basicBars"
output:
  html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We're going to create three different types of bar charts with the Elgin Response to Resistance data.

First we load the readr library and then the dataset.

```{r}
library(readr)
df <- read_csv("dfCrime.csv")
```

Now let's use summary() to remind us of what the data is

```{r}
summary(df)
```

Three years with four quarters each of response-to-resistance (RTR) incidents from the Elgin police department. That's broken down by the type of response - show of force only, use of force only and transition - show to use of force.

### Basic bar plot

Let's generate a bar plot of the total RTR incidents

```{r}
barplot(df$Total_RTR, 
        names.arg = df$Year_Quarter, cex.names=0.8,
        main="RTR incidents by year, quarter",
        xlab = "Number of incidents",
        col="blue")
```

Right off the bat, there's a problem. There aren't labels below for each bar, and which label applies to which bar isn't immediately clear.

If you're going to do graphics, not just for yourself but others, you're going to have to pay attention to these kinds of aesthetics. 

There's a lot of ways to address this, but the easiest is by making this a horizontal bar chart. We do this by adding the command __horiz=TRUE__

```{r}
barplot(df$Total_RTR, 
        horiz=TRUE,
        names.arg = df$Year_Quarter, cex.names=0.8,
        main="RTR incidents by year, quarter",
        xlab = "Number of incidents",
        col="blue")
```

And that's worse. We need to change the parameters of our plots so the axis text is horizontal too. We do that using the par command.

For more information on par, use the help method ?par

```{r}
par(las=1)

barplot(df$Total_RTR, 
        horiz=TRUE,
        names.arg = df$Year_Quarter, cex.names=0.8,
        main="RTR incidents by year, quarter",
        xlab = "Number of incidents",
        col="blue")
```

__barplot(__ is the command to make bar plots

__df\$Total_RTR__ is the column we want to plot

__horiz=TRUE__ creates a horizontal plot

__names.arg__ are the labels for the bars, __cex__ makes them a bit smaller.

__main__ and __xlab__ are obvious. __col__ is the color to use for the bars.

Sorting is automatic. There are ways to sort the data differently in a basic plot like this differently, but they're very complicated.

###Stacked bar plot

In addition to the total RTR, we have each quarter broken down by the type of response. A stacked bar plot would be very useful for this.

But first, we have to load only the data we want into a matrix and then transpose it. 

```{r}
counts2 <- matrix(c(df$SOF_only,df$UOF_only,df$Transitions),ncol=3)
colnames(counts2)=c("SOF_only","UOF_only","Transitions")
rownames(counts2)=df$Year_Quarter
counts2
```

__counts2 <- matrix(__ takes the list of the columns we want, creates a matrix and puts them into __counts2__. __ncol=3__ specifies the number of columns in the matrix.

__colnames(counts2)__ assigns names to the columns

__rownames(counts2)__ assigns names to the rows from the Year_Quarter column.

Now we need to transpose it.

```{r}
counts <- t(counts2)
counts
```

Compare counts2 to counts to see the difference.

Now we can plot

```{r}
par(las=1)

barplot(counts, horiz=TRUE, cex.names=0.8,
        main="RTR incidents by year, quarter",
        xlab = "Number of incidents",
        col=c("green","blue","red"),
        legend = rownames(counts),
        args.legend = list(x ='bottomright', inset=0.01)
)
```

__barplot__ needs the __par__ command to turn the axis labels. Generally you only need this once in a file.

__legend__ adds a legend based on the row names. Remember, we transposed the matrix so the columns became our rows.

__args.legend__ places the legend at the bottom right, __inset__ nudges it over a bit.

Try swapping in Counts2 instead of counts to see what happens.

### Grouped bar plot

The only difference between a stacked and a grouped bar plot is adding the __beside=TRUE__ command.

```{r}
par(las=1)

barplot(counts, horiz=TRUE, cex.names=0.8,
        beside=TRUE,
        main="RTR incidents by year, quarter",
        xlab = "Number of incidents",
        col=c("green","blue","red"),
        legend = rownames(counts),
        args.legend = list(x ='bottomright', inset=0.01)
)
```

Aesthetically, you need to decide which is better - stacked or grouped. For this data, stacked is more appropriate since the data add up to a total. 

But if you're comparing two completely different measures over time, then grouped would be best.